Lowestcase and Uppestcase letters: Further adventures in Derp Learning

### 2021-03-12 trunk/lowercase/paper/lowercase.tex
Hallucinate uses cache (huge speedup) and outputs 64 bit mask in bitmap mode. Generate (traced) bitmap SFDs. Little bit in the paper

@@

We want to machine-learn two functions, \makelowercase\ and
\makeuppercase. Each takes a letterform and returns a letterform (we
can choose how these are represented) and does the needful, e.g.
\makelowercase(\letterform{A}) should return \letterform{a}. In order to learn this function, we'll at least need a lot of
examples to use as training data. A training example for
\makelowercase\ is a letterform and its expected corresponding
lowercase one. We can ``easily'' find a large amount of examples by
using existing fonts, and pairing their \letterform{A} with their
\letterform{a}, and so on for all 26 letters, and symmetrically for
\makeuppercase.

@@

Sprechen of fonts, I downloaded every font I could find on the whole
internet. This was overkill. The resulting directory tree contained
over 100,000 files, many of which were duplicates. Exact duplicates
are easy to find, but since many of these files were the result of 30
years of community transmission, they had acquired various mutations.
One of the first things I did was write software to automatically
remove files that were essentially duplicates even if they weren't
exactly the same bytes.

@@

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{casealignment}
\caption{ Example alignment to reject the font
  {\tt DrippingGooExtended}.
  At left, \letterform{A} (red) and
  \letterform{a} (green) rendered with the identity transform, and
  their alignment ($35\%$ difference) below. At right, the transform
  found by the black-box optimizer and the resulting alignment with
  $1.7\%$ difference. Note that the shapes are still not an exact
  match (the export process has to round the data to integers and
  might apply other non-linear transformations like curve
  simplification), but these are clearly not a useful pair for
  the current problem.} \label{fig:casealignment}
\end{figure}

@@

\section{The simplest thing that might work}

Before getting fancy (which we we will) it's good engineering hygiene
to try the simpliest thing that might just work (it doesn't). Fonts
are represented as vector data (lines and quadratic B\'ezier curves).
Can we just train a network that takes these lines and curves as input
and predicts the lower- or uppercased letter in the same format? (No.)

@@

% XXX total number of parameters?
The network I trained has an input layer size of $154 = (38 + 14 + 10)
\times 4 + 3 \times 2$ (one control point and one end point per
B\'ezier curve), plus a starting point for each of the three contours.
The output layer is the same size, plus 26 (see below). There are
three hidden layers of size 308, 308, 360. The first two layers are
dense and the third sparse (about half of the weights are zero). All
layers are leaky rectified linear units. If you're taking notes,
don't, as again this does not work well, and I don't know how people
figure out what the right network size is anyway. I just made it up.
You can give {\em me} your notes.

@@

% TODO: Screenshot of this thing's output.
% TODO: Font file generated from this.
% TODO: Fitting the prediction to the actual in order to coax
% better results, failing.

@@

@@

% Perfect letters, hallucinated

>>>>

\renewcommand\paragraph[1]{\smallskip{\bf #1}\enspace}

@@

We want to machine-learn~\cite{neuralnetwork} two functions, \makelowercase\ and
\makeuppercase. Each takes a letterform and returns a letterform (we
can choose how these are represented) and does the needful, e.g.
\makelowercase(\letterform{A}) should return \letterform{a}. In order to learn this function, we'll at least need a lot of
examples to use as training data. A training example for
\makelowercase\ is a letterform and its expected corresponding
lowercase one. We can ``easily'' find a large amount of examples by
using existing fonts, and pairing their \letterform{A} with their
\letterform{a}, and so on for all 26 letters, and symmetrically for
\makeuppercase.

@@

Sprechen of Fonts, I downloaded every font I could find on the whole
internet. This was overkill. The resulting directory tree contained
over 100,000 files, many of which were duplicates. Exact duplicates
are easy to find, but since many of these files were the result of 30
years of community transmission, they had acquired various mutations.
One of the first things I did was write software to automatically
remove files that were essentially duplicates even if they weren't
exactly the same bytes.

@@

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{casealignment}
\caption{ Example alignment to reject the font {\tt
    DrippingGooExtended}. At left, \letterform{A} (red) and
  \letterform{a} (green) rendered with the identity transform, and
  their alignment ($35\%$ difference) below. At right, the transform
  found by the black-box optimizer and the resulting alignment with
  $1.7\%$ difference. Note that the shapes are still not an exact
  match (probably errors introduced in the font export process, which
  has to round the data to integers and might apply other non-linear
  transformations like curve simplification), but these are clearly
  not a useful pair for the current
  problem.} \label{fig:casealignment}
\end{figure}

@@

\section{The simplest thing that might work} \label{sec:vectorversion}

Before getting fancy (which we we will) it's good engineering hygiene
to try the simplest thing that might just work (it doesn't). Fonts
are represented as vector data (lines and quadratic B\'ezier curves).
Can we just train a network that takes these lines and curves as input
and predicts the lower- or uppercased letter in the same format? (No.)

@@

The network I trained has an input layer size of $154 = (38 + 14 + 10)
\times 4 + 3 \times 2$ (one control point and one end point per
B\'ezier curve), plus a starting point for each of the three contours.
The output layer is the same size, plus 26 (see below). There are
three hidden layers of size 308, 308, 360. The first layer is dense
and the remainder are sparse,
%(two thirds to a half of the weights are zero),
for just about 1 million total parameters. All layers are leaky
rectified linear units. If you're taking notes, don't, as again this
does not work well, and I don't know how people figure out what the
right network size is anyway. I just made it up. You can give {\em me}
your notes.

\paragraph{Bonus outputs.} The output includes the predicted shape,
and also 26 individual predictors for each of the 26 letters. So a
training example is actually like \letterform{C} $\rightarrow$
\letterform{c} $[0, 0, 1, 0, 0, \ldots, 0]$, with the 1 in the third
place because C is the third letter. We don't need these outputs for
the problem itself (e.g. to lowercase new letter shapes), but there
are several ideas behind this. First, the lowercasing function we're
trying to learn does take into account the letter of the alphabet
being lowercased. By asking the network to learn this as well (it is
penalized when it gets the prediction wrong), it must learn features
that allow it to distinguish different letters, and those features are
available for use by outputs we {\em do} care about. This is an easy
way to coax it to learn features that I know are meaningful without
having to actually engineer feature extractors by hand (or first train
separate models, etc.). Perhaps the most useful thing is that it's
very clear what the right answer is, so it gives me an easy way to see
if the network is learning anything {\em at all}. (It does.) Finally,
we can do some silly stuff with these; see
Section~\ref{sec:hallucination}.

@@

\begin{figure}[ht]
\includegraphics[width=0.9 \linewidth]{trainingvector}
\caption{ Screenshot of {\tt train.exe} running on the vector-based
  version of the problem. Shown is the \makelowercase\ model's output
  (bottom) on four shapes from four fonts (top). Some dust in between
  is the activation of the network's layers. At the very bottom, the
  26 predictions for ``what letter is this?''. The output for
  \letterform{j} is not too bad; you can see the distinct dot (a
  separate contour) and it sort of looks like a \letterform{j}. The
  \letterform{e} also has two pieces as expected but is otherwise
  garbage. The model is unsure whether the second input is an H
  or a K, and has predicted a shape sort of ambiguously between those
  two. The \lowercase{z} is also an embarrassment.
} \label{fig:trainingvector}
\end{figure}

I was so {\tiny confident} that this wouldn't work that I only trained
a \makelowercase\ model and didn't even worry about the more
complicated simultaneous training setup yet. I ran this for about
22,000 rounds, some 90 million training examples. Indeed it does not
work (Figure~\ref{fig:trainingvector}). It is not a total catastrophe,
though. We can tell from the 26 bonus outputs that the model can
clearly recognize letters (though perhaps just by memorization). Some
of the shapes it generates are along the right lines. ({\em Along the
  right lines}, get it??) I did not feel ANGRY at these results because
I expected it to not really work.

\paragraph{Just try making it more complicated!} This
problem of predicting the vector shape directly is a lot to ask of a
neural network, at least set up this way. One thing that did not sit
well with me is that the network could generate a perfect-looking
result, but because it didn't have the points in the expected order,
it would be penalized. This makes it harder to learn. True, but
my reflex to make things more complicated XXX

Ideally we would compute the difference between the predicted shape
and the expected shape, regardless of how they're drawn. Aside from
being a bit computationally challenging, this won't really work
because we need to attribute error directly to each output in order to
actually update the model in training. So I spent an afternoon writing
a routine to compute the best alignment of points between the actual
and expected outputs (Figure~\ref{fig:rotatej}). Aside from being
harder than it looked, my alignment code ended up being pretty slow
relative to the rest of training, even worse since it ran on the CPU
instead of GPU, and the training speed became UPSETTINGLY SLOW. So, I
discarded this excursion and threw this whole endeavour in the trash!
But I must have confused the Recycle~Bin icon with the fairly
complicated export-to-Truetype Font process that I built, because I
ran the model on the venerable Futura~\cite{futura} font and generated
{\bf Futurda} (Figure~\ref{fig:futurda}), which can be downloaded from my
website and installed for your desktop publishing needs.\footnote{Font
  downloads are available at \url{http://tom7.org/lowercase/}.}

% TODO: actually upload the fonts and make sure the URL is what you want

\begin{figure}[ht]
\includegraphics[width=0.5 \linewidth]{rotatej}
\caption{ Example alignment (from debug UI) of predicted shape (blue)
  to expected shape (green). We require each point to be mapped (red)
  to a point from the expected contour in a monotonic order (but
  several can be mapped to the same one), so that we can attribute
  error to each point. I hope it was fun to write this code because
  it was too slow to do pretty much anything but generate this figure!
} \label{fig:rotatej}
\end{figure}

% XXX add downloadable font too
\begin{figure*}[tp]
\includegraphics[width=0.9 \linewidth]{futurda}
\caption{ Type specimen for the generated font {\bf Futurda}. This is the
  classic font Futura, run through the vector based lowercasing model
  (Section~\ref{sec:vectorversion}). At top are Futura's letterforms
  \letterform{A}--\letterform{Z} run through the model and so ``made
  lowercase'' (it's obviously garbage). Next are
  \letterform{a}--\letterform{z},
  made even more lowercase. Also rubbish.
  At the bottom are the pangrams ``Dr.~Jock, TV Quiz Ph.D., bags few lynx''
  and ``Sphinx of black quartz, judge my vow!'' Although the text is
  illegible and kind of uninteresting, it does have a certain dynamic
  Rorschach aesthetic, like a collection of delicate moths pinned
  to paperboard, that one could consider framing or publishing in
  the proceedings of SIGBOVIK 2021.
} \label{fig:futurda}
\end{figure*}

% lowerercase Franklin Gothic. Franklin mint... not guaranteed
% to go up in value

% obviously, need to express different cases of anger at various
% problems or successes?

@@

\section{SDFs}

Don't give up! The fixed-size input/output of neural networks is
better suited to something like an array of pixels, and fonts can of
course be represented this way as well. To stay in the realm of what
my desktop computer with a single GeForce 1080 can do, I wanted to
keep the number of inputs and outputs pretty small. There's already an
excellent technique for representing font data as compact bitmaps,
which comes from computer graphics, called Signed Distance Fields
(SDFs)~\cite{green2007improved}. In a standard rasterization of a
font, each pixel of a bitmap contains 1 or 0, or perhaps an
anti-aliased value in-between. In an SDF, the bitmap instead contains
the distance to the nearest edge, and is signed (e.g. values inside
the shape are $> 0$, value outside are $< 0$). Actually in
practice we offset and saturate the values so that they are all in
$[0,1]$ (or bytes in $[0,255]$), with some nonzero ``on-edge value''
(say, 0.5) standing for ``distance 0''. In order to display the font
at the size of your choice, you then resize the SDF image with bilinear
interpolation, and then simply threshold the image. This works
surprisingly well (Figure~\ref{fig:sdf}).

SDFs seem well-suited for machine learning. They contain more
information per pixel than a plain bitmap, so we can use a smaller
input and output size. On the input
side, extremal pixels that would almost never be set in a bitmap still
have significant information (distance to the character). The error
function is just pixel-by-pixel difference. The rendering of the
output is inherently tolerant of some noise because of the sampling
and thresholding. So, this seemed like it might work well! (It doesn't
work that well.)

I got some stats on the font database, and determined the following
parameters for the SDFs.
 (here, $36\times36$ pixel images)
% Figure: bitmap vs SDF

@@

\section{Perfect letters, hallucinated}
\label{sec:hallucination}
